A weakly structured stem for human origins in Africa

### 2022-02-18 paper.tex
fix up author block
FIgure 1 legend

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[4]{Marlo M\"{o}ller}
\author[2,6,^,*]{Brenna M. Henn}
\author[7,^,**]{Simon Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, TX, USA}
\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[^]{Co-Corresponding Authors}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Geography and genetic diversity within Africa.}
        (A) The combined 1000 Genomes and AGRP datasets include populations from
        West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
        (D) illustrate the well-characterized high levels of genetic diversity
        and signatures of recent migration and gene flow within Africa.
        We built parameterized models using the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual (bold fonts).
        \aprcomment{work on this caption}
    }
    \label{fig:1}
\end{figure}

@@

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (see below). Fixed parameters included
divergence time the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). Such constraints, informed by prior research,
allowed us to infer robust migration rates. For example all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4}$). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
\citep{Alexander2009-sw} (Figure~\ref{fig:1}).

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fig2}
    \caption{
        \textbf{Two best fit models with a weakly structured stem.}
        In both the the parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are incidated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}), and which lasts for the
        duration of concurrence of contemporaneous populations. The
        merger-with-stem-migration model (B) outperformed the
        continuous-migration model (A), $LL=-102,600$ vs $-115,500$,
        respectively.
    }
    \label{fig:2}
\end{figure}

>>>>

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[4]{Marlo M\"{o}ller}
\author[2,6,$\dag$,*]{Brenna M. Henn}
\author[7,$\dag$,**]{Simon Gravel}
\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, TX, USA}
\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
\affil[$\dag$]{Co-Corresponding Authors}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig1.pdf}
    \caption{
        \textbf{Representative genetic diversity within Africa.}
        (A) Select populations from 1000 Genomes and ADRP illustrate diversity from
        western, eastern and southern Africa. We chose representatives from each region
        to build parameterized models, including the newly-sequenced Nama from South Africa, Mende
        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
        British and Vindija Neanderthal individual (bold fonts). (B, C) PCA highlights the range of genetic divergence anchored 
        by western Africans, Nama, Gumuz and the British. (D) \texttt{ADMIXTURE}
        illustrates signatures of recent gene flow in Africa which reflect colonial-period migration into the Nama,
        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
    }
    \label{fig:1}
\end{figure}

@@

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (see below). Fixed parameters included
the time of divergence between Western and Eastern African populations,
set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
We also fixed the amount of admixture from Neanderthals to Europeans directly
following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). Such constraints, informed by prior research,
allowed us to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
($m\approx2\times10^{-4}$). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/fig2}
    \caption{
        \textbf{Two best fit models with a weakly structured stem.}
        In both the the parameterizations of early population structure,
        continuous migration (A) and multiple mergers (B), models
        that include ongoing migration between stem populations outperform
        those in which stem populations are isolated. Most recent populations are
        connected by continuous, reciprocal migration that are indicated by 
        double-headed arrows (labels matched to migration rates and divergence
        times in Table~\ref{tab:migration-rates}), and which lasts for the
        duration of concurrence of contemporaneous populations. The
        merger-with-stem-migration model (B) outperformed the
        continuous-migration model (A), $LL=-102,600$ vs $-115,500$,
        respectively.
    }
    \label{fig:2}
\end{figure}
